---
title: "Chlorophyll Sandbox"
author: Document produced by C. Schlagenhauf, L. Dubroca, P. Derycke
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    number_section: yes
    theme: default
    toc: yes
    toc_depth: 2
    toc_float:
      collapsed: no
      smooth_scroll: yes
---
```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(
  eval = TRUE,
	echo = TRUE,
	fig.align = "center",
	message = FALSE,
	warning = FALSE,
	fig.width=5.5,
	out.width = "100%"
)
# clean environment
rm(list=ls())
gc()
```

```{r, include=FALSE, results='hide'}
# Chlorophyll Sandbox

Sessionid <- 'Chlorophyll Gradients'
source_provider <- "Copernicus Marine Service"
source_provider_url <- "https://marine.copernicus.eu"
layer_title<-"Chl-a gradients 2010-2021"
layer_Chla<-"Chlorophyll-a gradients 2010-2021"

# Define Region of interest (Bounding Box Longitude/latittude)
maxlat<- 51.4
maxlon<- 1.9
minlon<- -5.9
minlat<- 48

temp_path<- "/user/my-beach/"

xmin <- as.numeric(minlon)
ymin <- as.numeric(minlat)
xmax <- as.numeric(maxlon)
ymax <- as.numeric(maxlat)



```


The <b>Chlorophyll sandbox</b> is an analytical framework dedicated to the study of the Chlorophyll-A patterns. It leverages the scalability of an Elasticsearch cluster and enables a readily accessible digital twin ocean, fit for machine learning on big data. 20 years of satellite data at the global scale, more than 25 billion measurements, are available to deliver insights for any region of interest.

# Data information


Chl-a oceanic fronts are zones of rapidly changing phytoplankton and zooplankton biomass concentrations. These fronts, which can be observed by satellites (ocean colour), are structuring for a large number of natural processes like high biodiversity areas. They provide habitat for marine megafauna and have been proposed as proxies for pelagic diversity in the designation of marine protected areas. By extension, these fronts also support anthropic activities like fishing and aquaculture. The study of seasonal and interannual variation of fronts is important to better understand the physical, biogeochemical and ecological functioning of the marine environment.


<UL>
<LI>	Chlorophyll-a and SST monthly trends, anomalies and climatologies
<LI>	Environmental classification based on Chlorophyll-a and SST patterns
<LI>	Chlorophyll-a clustering analysis on 5 days datasets (interpolated with DIVA)
<LI>	Chlorophyll-a and SST forecast trends and gradients
</UL>

Data provider: https://marine.copernicus.eu


# Geographical extent


## Coordinates
```{r,echo=FALSE}
print (paste("West-Longitude:",round(xmin,2)))
print (paste("South-Latitude:",round(ymin,2)))
print (paste("East-Longitude:",round(xmax,2)))
print (paste("North-Latitude:",round(ymax,2)))
```






```{r, include=FALSE, results='hide'}

source_provider <- "Copernicus Marine Service"
source_provider_url <- "https://marine.copernicus.eu"

## First specify the packages of interest
packages = c("httr","rjson","ggplot2","raster","maptools","rasterVis","zoo","ggmap","dbscan")

## Now load or install&load all
package.check <- lapply(
  packages,
  FUN = function(x) {
    if (!require(x, character.only = TRUE)) {
      install.packages(x, dependencies = TRUE)
      library(x, character.only = TRUE)
    }
  }
)

```

```{r, include=FALSE, results='hide'}
## First specify the packages of interest
packages = c("wq")

## Now load or install&load all
package.check <- lapply(
  packages,
  FUN = function(x) {
    if (!require(x, character.only = TRUE)) {
      install.packages("./wq_0.4.8.tar.gz", repos=NULL)
      #package reshape2 needed
      #install.packages("'C:/temp/R/wq_0.4.8.tar.gz", repos=NULL)
      library(x, character.only = TRUE)
    }
  }
)

```

## Defined area

```{r ,echo=FALSE}

sr=SpatialPolygons(list(Polygons(list(Polygon(cbind(c(xmin, xmin, xmax, xmax),c(ymax, ymin, ymin, ymax)))),"1")))
mpa=SpatialPolygonsDataFrame(sr, data.frame(cbind(1:1), row.names=c("1")))
value<-(xmax-xmin)*(ymax-ymin)
if (value > 100) {
      zoom_value<-6
} else if (value > 1) {
      zoom_value<-7
} else {
      zoom_value<-8
}
base<-get_map(location=c(xmin-1,ymin-1,xmax+1,ymax+1), zoom=zoom_value, maptype="terrain-background", source = "stamen")
terrain <- ggmap(base)
map <- terrain + geom_polygon(data=mpa,aes(x=long,y=lat,group=group,fill="mpa"),colour="green",fill="blue",alpha=.1) +
ggtitle("")+xlab("Longitude")+ylab("Latitude")
plot(map)
```
Map tiles by <a href="http://stamen.com">Stamen Design</a>, under <a href="http://creativecommons.org/licenses/by/3.0">CC BY 3.0</a>. Data by <a href="http://openstreetmap.org">OpenStreetMap</a>, under <a href="http://www.openstreetmap.org/copyright">ODbL</a>.



```{r, include=FALSE, results='hide'}


size=10000

query<- paste0(' {
  "size": ',size,',
  "query": {
  "bool": {
  "must": {
  "match_all": {}
  },
  "filter": {
  "geo_bounding_box": {
  "geometry.coordinates": {
            "top_left": {
              "lat": ',ymax,',
              "lon": ',xmin,'
            },
            "bottom_right": {
              "lat": ',ymin,',
              "lon": ',xmax,'
            }
  }
  }
  }
  }
  }
  }
 ')

my_searchquery <- rjson::toJSON(query)

url_elk <- 'http://elk.knowcean.eu/elk/chla_'


i_stack=0
for (year in 2010:2021) {

month <- c("01","02","03","04","05","06","07","08","09","10","11","12")
for(month in month) {

data_list<-''
url_search <- paste0(url_elk,year,'-',month,'-01/_search')
print(url_search) 


response <- httr::POST(url = url_search,httr::add_headers(.headers = c("Content-Type" = "application/json")), body = fromJSON(my_searchquery), timeout(20))
data_list <- fromJSON(content(response, as="text"))

lat <- c()
lon <- c()
gradient <- c()
chla <- c()

for (i in 1:length(data_list$hits$hits)) {
  #print(data_list$hits$hits[[i]]$`_source`$gradient)
  gradient<-append(gradient,data_list$hits$hits[[i]]$`_source`$gradient)
  chla<-append(chla,data_list$hits$hits[[i]]$`_source`$chla)
  lat<-append(lat, data_list$hits$hits[[i]]$`_source`$geometry$coordinates[2])
  lon<- append(lon,data_list$hits$hits[[i]]$`_source`$geometry$coordinates[1])
}

myMatrix <- cbind(lon,lat,gradient)
myMatrixOrdered<-myMatrix[order(myMatrix[,1],decreasing=FALSE),]
dat <- as.data.frame(myMatrixOrdered)

myMatrix_chla <- cbind(lon,lat,chla)
myMatrixOrdered_chla<-myMatrix_chla[order(myMatrix_chla[,1],decreasing=FALSE),]
dat_chla <- as.data.frame(myMatrixOrdered_chla)


p1 <- ggplot(dat, aes(x = lon, y = lat, fill = gradient)) + geom_tile() + scale_fill_gradient2(low = "white", mid = "blue", high = "red3", midpoint = -1.0) + ggtitle(paste0("Year: ",year," Month: ",month))
plot(p1)


#create a raster with a spatial resolution close to the request
ncolras<-length(unique(dat$lon))
nrowras<-length(unique(dat$lat))
raster0<-raster(ncols=ncolras,nrows=nrowras,xmn=xmin,xmx=xmax,ymn=ymin,ymx=ymax)

#using rasterize
dat_raster<-rasterize(x=dat[,1:2],y=raster0,field=dat[,3],fun=mean,na.rm=T)
plot(dat_raster)
dat_raster_chla<-rasterize(x=dat_chla[,1:2],y=raster0,field=dat_chla[,3],fun=mean,na.rm=T)
plot(dat_raster_chla)

names(dat_raster) <- paste(year,"-",month, sep="")
names(dat_raster_chla) <- paste(year,"-",month, sep="")

if (i_stack == 0) {
imgs <- dat_raster
} else {
imgs <- raster::stack(imgs, dat_raster)
}

if (i_stack == 0) {
imgs_chla <- dat_raster_chla
} else {
imgs_chla <- raster::stack(imgs_chla, dat_raster_chla)
}

i_stack=i_stack+1

}
}


```

```{r, include=FALSE, results='hide'}


#FUNCTION PLOT PROCESS
#=================================================================
mpaprocessplot<-function (imgs, mpa, name, unite, logscale)
{
    if (dim(imgs)[3] < 12) {
        print("Series is too short (ie less than 12 months)")
    }
    else {
        imgsmean <- raster::mean(imgs, na.rm = T)
        imgsclim <- stackApply(imgs, indices = as.numeric(substr(names(imgs), 7, 8)), fun = mean, na.rm = TRUE)
        names(imgsclim) <- paste("Month", unique(as.numeric(substr(names(imgs), 7, 8))))
        titre <- paste(name, " (", unite, ")", sep = "")
pltall <- rasterVis::levelplot(imgs, zscaleLog = logscale, contour = T, col.regions = topo.colors(100), layout = c(12, NA), names = substr(names(imgs), 2, 8), main = titre)
        titre <- paste(name, " (", unite, ") ", substr(names(imgs)[1], 2, 8), "-", substr(names(imgs)[dim(imgs)[3]], 2, 8), " average", sep = "")
pltmean <- rasterVis::levelplot(imgsmean, margin = F, zscaleLog = logscale, contour = T, col.regions = topo.colors(100), main = titre)
        titre <- paste(name, " (", unite, ") monthly climatology", sep = "")
pltclim <- rasterVis::levelplot(imgsclim, zscaleLog = logscale, contour = T, col.regions = topo.colors(100), main = titre, names = names(imgsclim))
        titre <- paste(name, " (", unite, ") monthly boxplot", sep = "")
        return(list(pltall = pltall, pltmean = pltmean, pltclim = pltclim))
    }
}


#=================================================================
mpaprocessplotgradients<-function (imgs, mpa, name, unite, logscale)
{
    if (dim(imgs)[3] < 12) {
        print("Series is too short (ie less than 12 months)")
    }
    else {
        imgsmean <- raster::mean(imgs, na.rm = T)
        imgsclim <- stackApply(imgs, indices = as.numeric(substr(names(imgs), 7, 8)), fun = mean, na.rm = TRUE)
        names(imgsclim) <- paste("Month", unique(as.numeric(substr(names(imgs), 7, 8))))
        titre <- paste(name, " (", unite, ")", sep = "")
myPal <- RColorBrewer::brewer.pal('Blues', n=9)
myTheme <- rasterTheme(region = myPal)


# if one layer is only NA and logscale is TRUE then it bugs! Solution: replace NA by 0.1. The image is then monochrome
if (logscale == TRUE ){
imgs_pltall<-imgs
for (i in 1:dim(imgs_pltall)[3]){
if (is.null(imgs_pltall[[i]][values(imgs_pltall[[i]])!="NA"]) == TRUE){
imgs_pltall[[i]][is.na(imgs_pltall[[i]])] <- 0.1
}    
}
pltall <- rasterVis::levelplot(imgs_pltall, zscaleLog = logscale, contour = F, col.regions = topo.colors(100), layout = c(12, NA), names = substr(names(imgs), 2, 8), at = seq(0, mean(maxValue(imgsmean))*0.15, 0.05), main = titre)
        titre <- paste(name, " (", unite, ") ", substr(names(imgs)[1], 2, 8), "-", substr(names(imgs)[dim(imgs)[3]], 2, 8), " average", sep = "")
} else {
pltall <- rasterVis::levelplot(imgs, zscaleLog = logscale, contour = F, col.regions = topo.colors(100), layout = c(12, NA), names = substr(names(imgs), 2, 8), at = seq(0, mean(maxValue(imgsmean))*0.25, 0.05), main = titre)
}

titre <- paste(name, " (", unite, ") ", substr(names(imgs)[1], 2, 8), "-", substr(names(imgs)[dim(imgs)[3]], 2, 8), " average", sep = "")

pltmean <- rasterVis::levelplot(imgsmean, margin = F, zscaleLog = logscale, contour = T, col.regions = topo.colors(100), at = seq(0, mean(maxValue(imgsmean))*0.25, 0.05), main = titre)
        titre <- paste(name, " (", unite, ") monthly climatology", sep = "")
pltclim <- rasterVis::levelplot(imgsclim, zscaleLog = logscale, contour = T, col.regions = topo.colors(100), main = titre, at = seq(0, mean(maxValue(imgsmean))*0.2, 0.05), names = names(imgsclim))
        titre <- paste(name, " (", unite, ") monthly boxplot", sep = "")
        return(list(pltall = pltall, pltmean = pltmean, pltclim = pltclim))
    }
}





```


```{r, include=FALSE, results='hide'}

#processing of rasterstack containing the gradients/fronts data 

#Measurements are in log10. Convert to real values

imgs10<-10^imgs

#processing of data

plt_gradients10<-mpaprocessplotgradients(imgs=imgs10, mpa=mpa, name='Chl-a', unite='fronts', logscale=FALSE)


```


## Average fronts

```{r, echo=FALSE}

plt_gradients10[[2]]

```

## Monthly climatology

```{r, echo=FALSE}

plt_gradients10[[3]]

```

## Monthly composite

```{r, echo=FALSE}

plt_gradients10[[1]]

```

```{r, include=FALSE, results='hide'}
#FUNCTION DATA PROCESS
#=========================================================================
mpaprocessstat<-function (imgs, mpa, name, unite)
{
    datinpoly <- raster::extract(imgs, mpa, df = TRUE, cellnumbers = TRUE)
    tabdat <- as.vector(as.matrix(datinpoly[, 3:ncol(datinpoly)]))
    pixelnb <- nrow(datinpoly)
    tabts <- apply(datinpoly[, 3:ncol(datinpoly)], 2, mean, na.rm = T)
    fracnbna <- sum(is.na(tabts))/length(tabts)
    if (fracnbna > 0.2) {
        print("not enough data for time series analysis")
        vseason <- NA
        vtrend <- NA
        vtot <- NA
        testtrend <- data.frame(p.value = NA, sen.slope = NA)
        pltts <- NA
    }
    else {
        startyear <- as.numeric(substr(names(imgs)[1], 2, 5))
        startmonth <- as.numeric(substr(names(imgs)[1], 7, 8))
        tsm <- na.approx(ts(tabts, frequency = 12, start = c(startyear,
            startmonth)), na.rm = FALSE)
        tsmdec <- stl(tsm, s.window = 12, na.action = na.omit)
        tsseason <- tsmdec$time.series[, 1]
        tsremind <- tsm - tsseason
        linfit <- lm(tsremind ~ c(1:length(tsremind)))
        tstrend <- ts(predict(linfit), frequency = 12, start = c(startyear,
            startmonth))
        tsremind <- tsm - tsseason - tstrend
        pipo1 <- data.frame(time = as.numeric(time(tsm)), value = as.numeric(tsm),
            name = "Time series", type = "Original time series and trend")
        pipo2 <- data.frame(time = as.numeric(time(tsseason)),
            value = as.numeric(tsseason), name = "Seasonal component",
            type = "Seasonal and remainder components")
        pipo3 <- data.frame(time = as.numeric(time(tstrend)),
            value = as.numeric(tstrend), name = "Trend", type = "Original time series and trend")
        pipo4 <- data.frame(time = as.numeric(time(tsremind)),
            value = as.numeric(tsremind), name = "Residual",
            type = "Seasonal and residual components")
        pipoall <- rbind(pipo1, pipo2, pipo3, pipo4)
        pltts <- ggplot(pipoall, aes(x = time, y = value, group = name)) +
            geom_line() + facet_grid(name ~ ., scales = "free") +
            ylab(paste(name, " (", unite, ")", sep = ""))
        vseason <- var(tsseason, na.rm = T)
        vtrend <- var(tstrend, na.rm = T)
        vremind <- var(tsremind, na.rm = T)
        vtot <- vseason + vtrend + vremind
        testtrend <- mannKen(tstrend)
pipoa <- subset(pipo1, select=-c(name,type))
names(pipoa) <- c("time","Time series")
pipob <- subset(pipo2, select=-c(name,type))
names(pipob) <- c("time","Seasonal component")
pipoc <- subset(pipo3, select=-c(name,type))
names(pipoc) <- c("time","Trend")
pipod <- subset(pipo4, select=-c(name,type))
names(pipod) <- c("time","Residual")
pipox <- merge (pipoa, pipob, by.x="time", by.y="time" )
pipoy <- merge (pipoc, pipod, by.x="time", by.y="time" )
datatab <- merge (pipox, pipoy, by.x="time", by.y="time" )
    }
    datstat <- data.frame(name = name, unite = unite, timerange = paste(substr(names(imgs)[1],
        2, 8), substr(names(imgs)[dim(imgs)[3]], 2, 8), sep = "/"),
        pixelnb = pixelnb, fracnbna = round(sum(is.na(tabdat))/length(tabdat) *
            100, 2), fractsna = fracnbna * 100, mean = mean(tabdat,
            na.rm = T), sd = sd(tabdat, na.rm = T), min = min(tabdat,
            na.rm = T), max = max(tabdat, na.rm = T), varseason = 100 *
            vseason/vtot, vartrend = 100 * vtrend/vtot, trendtest = testtrend$p.value,
        sen.slope = testtrend$sen.slope)
    return(list(stat = datstat, pltts = pltts, datatab=datatab))
}

```


## Seasonal decomposition


```{r, include=FALSE, results='hide'}


#PROCESSING STAT SST
#========================================================================
datstat<-mpaprocessstat(imgs=imgs10,mpa=mpa,name='Gradients',unite='')
rezsst<-datstat[[1]]
datatabsst<-datstat[[3]]
names(datatabsst)[1]<-paste("Time")
#write.csv(datatabsst, file.path(temp_path,csvfile_name), row.names=FALSE, na="NaN")


```

```{r, echo=FALSE}

plot(datstat[[2]])

```

```{r, echo=FALSE}

datstat[[1]]


```


```{r, include=FALSE, results='hide'}

#FUNCTION PROCESS ANOMALIES
#========================================================================
anomalies<-function (imgs, mpa, name, unite, logscale)
{
    if (dim(imgs)[3] < 12) {
        print("Series is too short (ie less than 12 months)")
    }
    else {
                               titre <- paste(name, " (", unite, ")", sep = "")
                               pltall <- rasterVis::levelplot(imgs, zscaleLog = logscale, contour = F, col.regions = topo.colors(100), names = substr(names(imgs), 2, 8), main = titre)
                               imgsmean <- raster::mean(imgs, na.rm = T)
        titre <- paste(name, " (", unite, ") ", substr(names(imgs)[1], 2, 8), "-", substr(names(imgs)[dim(imgs)[3]], 2, 8), " average", sep = "")
                               pltmean <- rasterVis::levelplot(imgsmean, margin = F, zscaleLog = logscale, contour = F, col.regions = topo.colors(100), main = titre)
        imgsclim <- stackApply(imgs, indices = as.numeric(substr(names(imgs), 7, 8)), fun = mean, na.rm = TRUE)
        names(imgsclim) <- paste("Mean", unique(as.numeric(substr(names(imgs), 7, 8))))
        titre <- paste(name, " (", unite, ") monthly climatology", sep = "")
                               pltclim <- rasterVis::levelplot(imgsclim, zscaleLog = logscale, contour = T, col.regions = topo.colors(100), main = titre, names = names(imgsclim))
                               titre <- paste(name, " (", unite, ") monthly anomalies", sep = "")
                               imgsano<-imgs
                               for(i in 1:nlayers(imgsano)) {
                               j<-as.numeric(substr(names(imgsano[[i]]),7,8))
                               imgsano[[i]]<-imgsano[[i]]-imgsclim[[j]]
                               }
                               names(imgsano)<-names(imgs)
                               pltano <- rasterVis::levelplot(imgsano, zscaleLog = logscale, contour = F, par.settings = BuRdTheme, at = seq(-1.5, 1.5, 0.1), layout = c(12, NA), names = substr(names(imgsano), 2, 8), main = titre)
                               titre <- paste(name, " (%) monthly anomalies", sep = "")
                               imgsanopercent<-imgsano/imgs
                               names(imgsanopercent)<-names(imgs)
                               pltanopercent <- rasterVis::levelplot(imgsanopercent, zscaleLog = logscale, contour = T, par.settings = BuRdTheme, at = seq(-1.5, 1.5, 0.1), layout = c(12, NA), names = substr(names(imgsanopercent), 2, 8), main = titre)
                               return(list(pltall = pltall,  pltmean = pltmean, pltclim = pltclim, pltano = pltano, pltanopercent = pltanopercent))
    }
}



```

## Chlorophyll anomalies

```{r, include=FALSE, results='hide'}

aplt_chla<-anomalies(imgs=imgs_chla, mpa=mpa, name='Chl-a anomalies', unite='log10/milligram.m-3', FALSE)



```

```{r, echo=FALSE}

plot(aplt_chla[[4]])


```


## Front analysis

Here is an example of one month front analysis. I will be possible to count the fronts, 

```{r, include=FALSE, results='hide'}

## Select year and month

year<-"2021"
month<-"01"

url_search <- paste0(url_elk,year,'-',month,'-01/_search')
print(url_search) 


response <- httr::POST(url = url_search,httr::add_headers(.headers = c("Content-Type" = "application/json")), body = fromJSON(my_searchquery), timeout(20))
data_list <- fromJSON(content(response, as="text"))

lat <- c()
lon <- c()
gradient <- c()
chla <- c()

for (i in 1:length(data_list$hits$hits)) {
  #print(data_list$hits$hits[[i]]$`_source`$gradient)
  gradient<-append(gradient,data_list$hits$hits[[i]]$`_source`$gradient)
  chla<-append(chla,data_list$hits$hits[[i]]$`_source`$chla)
  lat<-append(lat, data_list$hits$hits[[i]]$`_source`$geometry$coordinates[2])
  lon<- append(lon,data_list$hits$hits[[i]]$`_source`$geometry$coordinates[1])
}

myMatrix <- cbind(lon,lat,gradient)
myMatrixOrdered<-myMatrix[order(myMatrix[,1],decreasing=FALSE),]
dat <- as.data.frame(myMatrixOrdered)

myMatrix_chla <- cbind(lon,lat,chla)
myMatrixOrdered_chla<-myMatrix_chla[order(myMatrix_chla[,1],decreasing=FALSE),]
dat_chla <- as.data.frame(myMatrixOrdered_chla)

```

### Map

```{r, echo=FALSE}

p2 <- ggplot(dat, aes(x = lon, y = lat, fill = gradient)) + geom_tile() + scale_fill_gradient2(low = "white", mid = "blue", high = "red3", midpoint = -1.0) + ggtitle(paste0("Year: ",year," Month: ",month))
plot(p2)


```

### Density

```{r, echo=FALSE}

#p1 <- ggplot(dat,aes(x=gradient,color=gradient,after_stat(count))) + geom_density() 
#plot(p1)
	
p1 <- ggplot(dat, aes(x = lon, y = lat, fill = gradient)) + geom_tile() + scale_fill_gradient2(low = "white", mid = "blue", high = "red3", midpoint = -1.0) + ggtitle(paste0("Log10 Year: ",year," Month: ",month))
plot(p1)

```



```{r, include=FALSE, results='hide'}

d<-dat[!(dat$gradient < -0.8),]

cl2 <- hdbscan(d, minPts = 10)
cl3<-fpc::dbscan(d$gradient, 0.1, MinPts = 8, scale = FALSE, method = c("hybrid"), seeds = FALSE, showplot = FALSE, countmode = NULL)

```

### Front classification


```{r, echo=FALSE}

plot(d, col=cl2$cluster+1, 
     pch=ifelse(cl2$cluster == 0, 8, 1), # Mark noise as star
     cex=ifelse(cl2$cluster == 0, 0.1, 0.01))

```

```{r, echo=FALSE}

plot(d$lon, d$lat, col=cl2$cluster+1, 
     pch=ifelse(cl2$cluster == 0, 8, 1), # Mark noise as star
     cex=ifelse(cl2$cluster == 0, 0.1, 0.01), # Decrease size of noise
     xlab=NA, ylab=NA)

```

```{r, echo=FALSE}

plot(d, col=cl3$cluster+1, 
     pch=ifelse(cl3$cluster == 0, 8, 1), # Mark noise as star
     cex=ifelse(cl3$cluster == 0, 0.1, 0.01))

```



